---
title: "R crash course / 05"
author: ""
date: "`r Sys.Date()`"
output:
  ioslides_presentation: default
---

```{r,echo=FALSE,warning=FALSE,message=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=TRUE,autodep=TRUE, results="hide")
require(readxl)
require(beeswarm)
require(tidyr)
require(vioplot)
require(RColorBrewer)
```

# SD, SEM and different plot types

## SD vs SEM


```{r,eval=TRUE}

## generate mock data sets with different sample sizes
ns <- c(10, 50, 100, 500)
data <- rnorm(sum(ns))
## repeat "a" 10 times, "b" 50 times, etc.
labs <- unlist(lapply(1:length(ns), 
  function(i) rep(letters[i], ns[i])))

## calculate means, SD and SEM
means <- tapply(data, labs, mean)
sds <- tapply(data, labs, sd)
ses <- sds/sqrt(ns)
```


## SD vs SEM (cont.)

```{r,echo=FALSE,eval=TRUE,fig.width=7,fig.height=5}
x <- (1:length(ns))-.3 # for plotting

plot(NULL, xlim=c(.5, length(ns) + .5), ylim=c(-5, 5), bty="n", xlab="", ylab="")
## means
abline(h=0, col="grey")
abline(h=c(-1,1), col="grey", lty=2)
segments(x, tapply(data, labs, mean), x+.6, tapply(data, labs, mean), lwd=2)

rect(x + .1, means-sds, x + .5, means + sds, col="#00999933")
rect(x, means-ses, x + .6, means + ses, col="#00FFFF33")
beeswarm(data ~ labs, pch=19, col="#33333333", add=TRUE)
```


## Function for plotting error bars

```{r eval=TRUE}
ebar <- function(x, y, sd, width=1, ...) {
  w <- width/2

  segments(x, y, x, y + sd, ...)
  segments(x - w, y + sd, x + w, y + sd, ...)
}
```


## SD vs SEM (cont.)


```{r eval=TRUE,echo=FALSE}
set.seed(121234)
data <- c(rnorm(100) + 1.6, rnorm(100) + 2)
labs <- rep(letters[1:2], each=100)
```

p < `r format.pval(t.test(data ~ labs)$p.value, digits=2)`

```{r eval=TRUE,echo=FALSE,fig.width=3,fig.height=5}
## calculate the means and SEMs
means <- tapply(data, labs, mean)
sems <- tapply(data, labs, sd)/sqrt(100)
bp <- barplot(means, ylim=c(0, 2.5), space=.9)
ebar(bp[,1], means, sems, width=.6)
```

## SD vs SEM (cont.)

```{r eval=TRUE,echo=FALSE,fig.width=8,fig.height=5}

layout(t(c(1,2)), widths=c(1,4))
## calculate the means and SEMs
bp <- barplot(means, ylim=c(0, 2.5), space=.1)
ebar(bp[,1], means, sems, width=.6)
bp <- barplot(means, ylim=c(0, 2.5), space=.6)
ebar(bp[,1], means, sems, width=.3)
```


## SD vs SEM (cont.)


```{r eval=TRUE,echo=FALSE,fig.width=8,fig.height=5}
par(mfrow=c(1,2))
bp <- barplot(means, ylim=c(0, 2.5), space=.9)
ebar(bp[,1], means, sems, width=.5)
boxplot(data ~ labs, frame=FALSE, notch=TRUE)
```


## SD vs SEM (cont.)

```{r eval=TRUE,echo=FALSE,fig.width=8,fig.height=5}
par(mfrow=c(1,2))
bp <- barplot(means, ylim=c(0, 2.5), space=.9)
ebar(bp[,1], means, sems, width=.5)
beeswarm(data ~ labs, bty="n")
```

## SD vs SEM (cont.)

```{r eval=TRUE,echo=FALSE,fig.width=8,fig.height=5}
par(mfrow=c(1,2))
bp <- barplot(means, ylim=c(0, 2.5), space=.9)
ebar(bp[,1], means, sems, width=.5)
vioplot(data ~ labs, col="#33993333", frame.plot=F)
```

# Linear regression

## Calculating linear regression

```{r,eval=TRUE}
mod1 <- lm(mpg ~ hp, data=mtcars)
summary(mod1)
mod2 <- lm(mpg ~ hp + disp, data=mtcars)
summary(mod2)
```

## Plotting linear regression: basic

Plotting mod2 would be less trivial, since we only show one predictor
  variable on the plot, but the model depends on two of them.

```{r eval=TRUE,fig.height=3,fig.width=4}
plot(mtcars$hp, mtcars$mpg)
abline(mod1)
```


## Comparing models 

```{r,eval=TRUE}
mod1 <- lm(mpg ~ hp, data=mtcars)
summary(mod1)
mod2 <- lm(mpg ~ hp + disp, data=mtcars)
summary(mod2)
anova(mod1, mod2)
mod.l <- lm(mpg ~ log(hp), data=mtcars)
BIC(mod1, mod2, mod.l)
```


## Plotting linear regression

Confidence intervals tell us how well we have estimated the relationship.
Prediction intervals tell us how well we can predict new data points using
this model.

```{r}
mod <- lm(mpg ~ log(hp), data=mtcars)
xnew <- 50:350
plot(mtcars$hp, mtcars$mpg, log="x", xlab="HP",
  ylab="MPG", bty="n")
ci <- predict(mod, newdata=data.frame(hp=xnew), interval="conf")
lines(xnew, ci[,1], lwd=2) # fit
lines(xnew, ci[,2]) # lower CI
lines(xnew, ci[,3]) # upper CI
pr <- predict(mod, newdata=data.frame(hp=xnew), interval="pred")
lines(xnew, pr[,2], lty=2) # lower prediction
lines(xnew, pr[,3], lty=2) # upper prediction
```


## Plotting linear regression (cont.)

```{r,eval=TRUE,echo=FALSE,fig.width=6,fig.height=5}
mod <- lm(mpg ~ log(hp), data=mtcars)
xnew <- 50:350
plot(mtcars$hp, mtcars$mpg, log="x", xlab="HP",
  ylab="MPG", bty="n")
ci <- predict(mod, newdata=data.frame(hp=xnew), interval="conf")
lines(xnew, ci[,1], lwd=2) # fit
lines(xnew, ci[,2]) # lower CI
lines(xnew, ci[,3]) # upper CI
pr <- predict(mod, newdata=data.frame(hp=xnew), interval="pred")
lines(xnew, pr[,2], lty=2) # lower prediction
lines(xnew, pr[,3], lty=2) # upper prediction
```


## Plotting linear regression cont.

```{r}
plot(NULL, xlim=c(50,350), ylim=c(5,45), log="x", xlab="HP",
  ylab="MPG", bty="n")
x2 <- c(xnew, rev(xnew))
polygon(x2, c(pr[,2], rev(pr[,3])), col="#FF330033", border=NA)
polygon(x2, c(ci[,2], rev(ci[,3])), col="#FF330033", border=NA)
lines(xnew, ci[,1], col="#FF3300", lwd=2) # fit
points(mtcars$hp, mtcars$mpg, pch=19, cex=1.3, col="#99110033")
```


## Plotting linear regression cont.

```{r,echo=FALSE,eval=TRUE,fig.width=7,fig.height=5}
plot(NULL, xlim=c(50,350), ylim=c(5,45), log="x", xlab="HP",
  ylab="MPG", bty="n")
x2 <- c(xnew, rev(xnew))
polygon(x2, c(pr[,2], rev(pr[,3])), col="#FF330033", border=NA)
polygon(x2, c(ci[,2], rev(ci[,3])), col="#FF330033", border=NA)
lines(xnew, ci[,1], col="#FF3300", lwd=2) # fit
points(mtcars$hp, mtcars$mpg, pch=19, cex=1.3, col="#991100")
```

# FC-FC plots, Disco

## Plotting fold change vs fold change

A good way to compare results of two analyses is to plot the log2 fold
changes vs each other.

```{r eval=TRUE}
res <- read.csv("Datasets/transcriptomics_results.csv")
```

##

```{r eval=TRUE,fig.width=5,fig.height=5}
plot(res$logFC.F.D1, res$logFC.F.D3, pch=19, col="#33333333", bty="n", 
  xlab="Fold change on D1 vs D0", ylab="Fold change on D3 vs D0")
abline(h=0)
abline(v=0)
```


## Disco score

```{r eval=TRUE}
library(disco)
disco <- list(
  genes=res$GeneName,
  lfc=res[ , c("logFC.F.D1", "logFC.F.D3")],
  pval=res[ , c("qval.F.D1", "qval.F.D3")])
disco$ds <- disco.score(disco)
plotDisco(disco, disco$ds)
```

## Disco score 2

```{r eval=FALSE}
disco2 <- data.frame(disco$lfc, ds=disco$ds)
disco2 <- disco2[ order(abs(disco2$ds)), ]
pal <- colorRampPalette(colors=c("blue", "lightgrey", "red"))(17)
pal <- paste0(pal, "66")
col <- tagcloud::smoothPalette(disco2$ds, pal=pal, min=-3, max=3)
plot(disco2[,1], disco2[,2], pch=19, bty="n", 
  xlab="Fold change on D1 vs D0", ylab="Fold change on D3 vs D0", col=col)
abline(h=0, col="grey")
abline(v=0, col="grey")
```

##

```{r eval=TRUE,echo=FALSE,fig.width=8,fig.height=6}
disco2 <- data.frame(disco$lfc, ds=disco$ds)
disco2 <- disco2[ order(abs(disco2$ds)), ]
pal <- colorRampPalette(colors=c("blue", "lightgrey", "red"))(17)
pal <- paste0(pal, "66")
col <- tagcloud::smoothPalette(disco2$ds, pal=pal, min=-3, max=3)
plot(disco2[,1], disco2[,2], pch=19, bty="n", 
  xlab="Fold change on D1 vs D0", ylab="Fold change on D3 vs D0", col=col)
abline(h=0, col="grey")
abline(v=0, col="grey")
```





# Heatmaps

## loading data

```{r eval=TRUE}
genes <- read_xlsx("Datasets/expression_data_vaccination_example.xlsx", 
  sheet=2)
e <- read_xlsx("Datasets/expression_data_vaccination_example.xlsx", 
  sheet=3)
t <- read_xlsx("Datasets/expression_data_vaccination_example.xlsx", 
  sheet=1)
res <- read.csv("Datasets/transcriptomics_results.csv")

## res and genes are not identical!
com <- intersect(genes$ProbeName, res$ProbeName)
e <- data.matrix(e[ match(com, genes$ProbeName), ])
genes <- genes[ match(com, genes$ProbeName), ]
rownames(e) <- genes$GeneName
res <- res[ match(com, res$ProbeName), ]
all(res$ProbeName == genes$ProbeName)
```


## Selecting genes to plot on a heatmap

```{r eval=TRUE}
n <- 25 # top 50
byqval <- order(res$qval.F.D1)
sel <- byqval[1:n]

downreg <- which(res$logFC.F.D1 < 0)
downreg <- byqval[ byqval %in% downreg ][1:n]

sel <- c(sel, downreg)


## ordering and selecting columns
group <- paste0(t$ARM, ".", t$Timepoint)
csel <- order(group)
## FLUAD only
csel <- csel[ csel %in% which(t$ARM == "FLUAD") ] 
```

## Plotting the heatmap

```{r}
library(gplots)
mtx <- t(e[sel, csel])

## scale manually
mtx <- scale(mtx)

## palette
pal <- colorRampPalette(c("cyan", "black", "purple"))
heatmap.2(mtx, trace="n", scale="n", dendrogram="c", 
  Rowv=FALSE, col=pal,
  breaks=seq(-2, 2, length.out=15) )
```


## 

```{r,eval=TRUE,echo=FALSE,fig.width=8,fig.height=6}
library(gplots)
mtx <- t(e[sel, csel])

## scale manually
mtx <- scale(mtx)

## palette
pal <- colorRampPalette(c("cyan", "black", "purple"))
heatmap.2(mtx, trace="n", scale="n", dendrogram="c", 
  Rowv=FALSE, col=pal,
  breaks=seq(-2, 2, length.out=15) )
```



# Simple clustering

## Hierarchical clustering

```{r eval=TRUE}
data(iris)
mtx <- data.matrix(iris[,1:4])
mtx <- scale(mtx)
group <- factor(iris[,5])

dd <- dist(mtx)
tree <- hclust(dd)
cl <- cutree(tree, k=3)
table(cl, group)
```

## K-means clustering

```{r eval=TRUE}
n.cl <- 3
km <- kmeans(mtx, n.cl)
table(km$cluster, group)
pal <- brewer.pal(n=n.cl, name="Dark2")
```

##

```{r eval=TRUE,fig.width=7,fig.height=6}
plot(mtx[,1], mtx[,2], col=pal[km$cluster], pch=c(15,17,19)[factor(iris[,5])], bty="n")
points(km$centers[,1], km$centers[,2], pch=19, cex=4, col="#99999966")
text(km$centers[,1], km$centers[,2], 1:n.cl, cex=2, col=pal)
legend("topright", levels(group), pch=c(15,17,19), cex=1.5, bty="n")
```




