---
title: "R crash course / 02"
author: ""
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    widescreen: true
    smaller: true
    css: files/style.css
    logo: files/logo_bihealth_en.png
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=TRUE,autodep=TRUE, results="hide")
```


## Repetitorium of day 1: Example

There is a lake in a garden. Every day, the water lilies cover twice as
much area as the previous day. On the first day, the water lilies cover
1/100th of the area of the lake.

 * What is the formula to calculate the area covered by the water lilies on
   day n? (pen and paper, not in Rstudio!)
 * assign days 1 ... 10 to variable x. Now assign the fraction of the area
   covered by water lilies on day n to variable y
 * plot x vs y (use the simple `plot()` function)
 * When will the half of the area be covered by water lilies? Use `abline(h=.5)`
   (what does it do?) to show a graphical solution
 * What is the fraction on day 3? Hint: what does `y[3]` do?
 * Speaking of which, what dows `y[4:5]` do? (just try it!)

## Exercise

 * Make the plot yourself.
 * In Rmarkdown.
 * use 'col' parameter of `plot` and a color name (e.g. "red") to change
   the color of the line (`plot(..., col="red")`)
 * what if each day the lilies cover a fraction of area that is 1.5 the
   fraction of area on the previous day?
 * Use `lines(x, y)` function to put a second line on the plot


## Creating your own functions for fun and profit

 * If you have to write something more than twice, you should turn it into
   a function!
 * R is a *functional* language

## Creating a function

```{r eval=FALSE,echo=TRUE}
#' Function name
#' Function description
some_name <- function(param1, param2=2) {

## code comment
# <your code goes in here>

}
```

## Example

```{r eval=FALSE,echo=TRUE}

#' Function for pond calculation
pond_surface_covered <- function(init_pond_surface, day=0) {

  ## "**" and "^" do the same thing
  ret <- init_pond_surface * 2^day

  return(ret)
}
```


## Example


```{r eval=FALSE,echo=TRUE}

#' Generate a random sample from a given distribution
#' Usage: randsamp(n) where n is the size of the sample

randsamp <- function(n, d="uniform") {

  if(d == "uniform") {
    ret <- runif(n)
  } else if(d == "normal") {
    ret <- rnorm(n)
  } else {
    stop("I don't know this distribution!")
  }

  return(ret)
}
```

## Exercise 2.1

Take the function above and add a new option: the binomial distribution.
Which function can give you that? Search for a suitable function.

Note: binomial distribution is essentially what you get when you throw a
coin a number of times and add the results (1 for heads, 0 for tails). 
This function requires two more parameters: number of "coin tosses"
and probability of "heads". You need to adapt the code accordingly: choose
something meaningful, e.g. size=1 and prob=0.5 to simulate a single fair
coin toss.



## Data matrices

Much like vectors, matrices can only hold one data type (e.g. only numeric
or only character or only logical etc.).

```{r}
m <- matrix(1:18, ncol=3, nrow=6)
# compare with
m <- matrix(1:18, ncol=3, nrow=6, byrow=TRUE)
dim(m)
ncol(m)
nrow(m)
```

## Accessing columns, rows and element

`matrix[row, column]`

So, for example:

```{r}
m[1, ] # vector which is the first row
m[, 2] # vector which is the first column
m[3, 1] # first element of the third row
```

## Row and column names

We can name rows and columns of a matrix and use the names to access the
rows and columns:

```{r}
colnames(m) <- letters[1:ncol(m)]
rownames(m) <- LETTERS[1:nrow(m)]

m["A", "b"]
m["B", ] # etc.
```

## Remember

 * If you select a single column or a single row, you will get a vector
 * If you select more than one row or column, you will get a (smaller) matrix
 * If you select more rows or columns than are present, you will get a
   "subscript out of bonds" error

## Demonstration


## Lists

 * Lists hold all types of information
 * Lists are really, really cool
 * Lists have elements. An element of a list can have *any* type, including
   another list.

## Lists

 Caveats:

 * You access elements of a list using `[[`, not `[`
 * Lists may have names (set with `names()`), but don't have to

## Data frames

Data frames are a bit like matrices, but every column can store different
type of data. In this, they are more like lists (which they in fact are).

```{r}
names <- c("January", "Bill")
lastn <- c("Weiner", "Gates")
age   <- c(1001, 64)

d <- data.frame(names=names, last_names=lastn, age=age)
class(d)
class(d[,1])
class(d[,3])
```

## Accessing elements in data frames

You can access the data frame elements much like the elements of a matrix.

However, since data frames are lists, the list operator (`$`) also works:

```{r}
d$names # same as d[,1] or d[, "names"]
d$lastn
d$lastn[1]
```

However, note that when you select a row, you will get a data frame, not a
vector. This is because each of the column can be of different type, and
vectors can hold only one type of data.

## Matrices vs data frames

 Caveats:

  * data frames sometimes turn strings to factors (more on that later), which may have really
    disastrous consequences, use `stringsAsFactors=FALSE`
  * small typos can turn a numeric vector into a string
  * factors are dangerous to work with, use them cautiously

  Gory details: matrices are a basic data type. Data frames are a list.

## Data frames vs tibble

 Caveats:

 * tibbles are the data frames from tidyverse
 * Whatever you can do to a data frame, you can do to a tibble as well
 * `read_*` functions return a tibble
 * tibble do not have row names
 * If you select a single row in a data frame, you get a smaller data
   frame. If you select a single column, you get a vector.
 * In tibble, you always get a smaller tibble.


## Exercise 2.1

 * Create a 5x3 matrix with random numbers. Use `matrix` and `rnorm`.
 * Turn the matrix into a data frame. Use `as.data.frame` for that.
 * Add column and row names.
 * Add a column. Each value in the column should be "A" (a string). Use the
   `rep` function for that.
 * Add a column with five numbers from 0 to 1. Use the `seq` function for that.

## Using rep

`rep()` is used to replicate vectors.

```{r}
rep(c("A", "B"), 5) 
# result:
#  [1] "A" "B" "A" "B" "A" "B" "A" "B" "A" "B"

rep(c("A", "B"), each=5)
# result
# [1] "A" "A" "A" "A" "A" "B" "B" "B" "B" "B"
```

# Reading and writing data

## Reading data

Main data types you will encounter:

---------------------------- ------------------------------- --------------------------
Data type                    Function                        Notes
---------------------------- ------------------------------- --------------------------
TSV / TAB separated values   `read_table()`                  Tab is `\t`.
                                                             A very general and 
                                                             customizable function

CSV / comma separated        `read_csv()`                    a wrapper around
                                                             `read.table()`

XLS (old Excel)              `read_xls()`                    Just don't use it.
                             `read_excel()`                  From the
                                                             `readxl`
                                                             package.

XLSX (new Excel)             `read_xlsx()`                   From the `readxl` package.
                             `read_excel()`                  You need to
                                                             provide the
                                                             sheet number
                                                             you wish to
                                                             read. 
                                                             Note: returns a tibble,
                                                             not a data frame! 
---------------------------- ------------------------------- --------------------------



## Excercise 2.2

Read, inspect and correct (if reasonable) the following files:

 * `cars.xlsx`
 * `iris.csv`
 * `expression_data_vaccination_example.xlsx` (large!)

The function `readxl_example("deaths.xls")` returns a file name. Read this
file. How can you omit the lines at the top and at the bottom of the file?
(hint: `?read_xls`). How can you force the date columns to be interpreted
as dates and not numbers?


## Tibbles / readxl

tibbles belong to the tidyverse. They are nice to work with and very
useful, but we can stick to data frames for now. Therefore, do

```{r}
mydataframe <- as.data.frame(read_xlsx("file.xlsx"))
```

One crucial difference between tibble and data frame is that `tibble[ , 1 ]` 
returns a tibble, while `dataframe[ , 1]` returns a vector. The second
crucial difference is that it does not support row names (on purpose!).

## Cleaning up data

`unique` returns the unique values in a vector. Use it to check the
correctness of the columns.

Use `class()` to make sure that the columns are what you think they are.

Use `summary()` to gain an overview of the data.

Convert with `factor` when necessary (and only then).

## Substitutions (search & replace)

 * `gsub(string1, string2, text)` substitute all occurences of `string1` in
    by `string2` in `text`
 * `sub(...)` same, but only the first occurence

```{r}
## Clearly broken!
group <- c("Control", " control", "control ", "Control   ")
group <- gsub("Control", "control", group)
group <- gsub(" ", "", group)
```

## Regular expressions

Is there not a way to simplify things? 

```{r}
group <- c("Control", " control", "control ", "Control   ")

# one method
gsub(" ", "", group) # but what happens if the groups are "group 1", "group 2" etc?

# substitute at the end of the string
gsub(" $", "", group)

# at the beginning
gsub("^ ", "", group)

# any number of occurences
gsub(" *", "", group)
```

## Regular expression cont.

In regular expressions, some characters or strings have special meanings

 * Some mean what is supposed to be replaced (e.g. `.`, dot,
   means "any character")
 * Some mean how many times something gets matched; e.g. `*` means "any
   number of matches of the previous character, so `.*` means any character
   any number of times (like, everything)
 * Some characters at certain position may also have a special meaning
   (like `^` means "match the beginning of a string" and `$` at the end of
   a regular expression means "match at the end of a string").
 * A string like `[abc]` will match only a, b and c, but not d. You can use
   ranges, e.g. `[0-9]` or `[a-z]`

## Regular expression cont.

```{r}
foo <- c("a1_123", "a1_231", "a2_17", "B2_23", "c2_889")

# substitute digits only
gsub("[0-9]*", "", foo)

# substitute letters only
gsub("[a-z]*", "", foo)

# same, but case insensitive
gsub("[a-z]*", "", foo, ignore.case=TRUE)

# substitute everything after "_"
# dot means "any character"
gsub("_.*", "", foo)
```

## Regexps cont.

You can wrap gsub around another gsub

```{r}
vec <- c("mouse", " Mouse", "mus domesticus", "chicken", "Schicken", "Duck")
vec <- gsub("S*chicken", "Chicken", gsub("^m.*", "Mouse", gsub("^ *", "", vec)))

## same as

vec <- gsub("^ *", "", vec)
vec <- gsub("^m.*", "Mouse", vec)
vec <- gsub("S*chicken", "Chicken", vec)
```

## Exercice 2.regexp

Figure out how to use gsub to do the following tasks:

 * What matches all of the "ANKRD" genes in the following: `c("ANKRD22", "ank.rep.domain 12", "ifng-1", "ANKRD-33", "  ankrd23")`
 * Make the following uniform: `c("male", "Male ", "M", "F", "female", " Female")`
 * Using `gsub` and the `toupper` function, clean up the gene names such
   that they conform to the HGNC (all capital
   letters, no spaces, no dashes): `c("ankrd22", "ifng", "Nf-kb", " Cxcl 5", "CCL 6.", "ANK.r.d. 12")`
 * Read the data file `meta_data_botched.xlsx`. Spot and correct errors.

## Writing data

Main data types you will encounter:

---------------------------- ------------------------------- --------------------------
Data type                    Function                        Notes
---------------------------- ------------------------------- --------------------------
TSV / TAB separated values   `write.table()`                 Tab is `\t`.
                                                             A very general and 
                                                             customizable function
                                                             Warning:
                                                             column names 

CSV / comma separated        `write.csv()`                    a wrapper around
                                                             `write.table()`

XLS (old Excel)                                              Just don't use it. No, 
                                                             seriously, don't.

XLSX (new Excel)             `write_xlsx()`                  From the `writexl` 
                                                             package.
                                                             Warning:
                                                             row names
                                                             omitted!
---------------------------- ------------------------------- --------------------------



## Wide and Long format (demonstration)


## Wide and Long format

 Long advantages:

  * easier to filter, process, visualize, do statistics with
  * focused on measurement ("patient ID" or equivalent is a covariate, and so is measurement type)
 
 Wide advantages:

  * groups data by a covariate ("patient ID")
  * can be easier to manage (each column one measurement type)



## Converting from wide to long:

```{r}
wide <- read.table(header=TRUE, text='
 subject sex control cond1 cond2
       1   M     7.9  12.3  10.7
       2   F     6.3  10.6  11.1
       3   F     9.5  13.1  13.8
       4   M    11.5  13.4  12.9
')
pivot_longer(wide, cols=c("control", "cond1", "cond2"), 
  names_to="condition", values_to="measurement")
```



## Converting from long to wide

```{r}
long <- read.table(header=TRUE, text='
 subject  sampleID sex condition measurement
       1  ID000001 M   control         7.9
       1  ID000002 M     cond1        12.3
       1  ID000003 M     cond2        10.7
       2  ID000004 F   control         6.3
       2  ID000005 F     cond1        10.6
       2  ID000006 F     cond2        11.1
       3  ID000007 F   control         9.5
       3  ID000008 F     cond1        13.1
       3  ID000009 F     cond2        13.8
')
```

## Converting from long to wide

```{r}
## not what we wanted!!! Why?
pivot_wider(long, names_from="condition", values_from="measurement")

## Instead: 
pivot_wider(long, id_cols="subject", names_from="condition", values_from="measurement")
```

## Exercise 2.3

Convert the following files to long format:

 * `iris.csv`
 * `mtcars_wide.csv`
 * `labresults_wide.csv`
 * `cars.xlsx` (tricky!)

## Exercise 2.4

First: how do you specify the sheet of an Excel file with `read_xlsx`? How
do you find out what sheets are in the file?

For the first sheet ("targets") from the file
`expression_data_vaccination_example.xlsx`, create a new table. For each
Patient ID there should be one row. The row should contain a column for
each day (D0, D1 etc.) containing the sample IDs collected on that day.

## A short note on organizing your data

 * If possible, use a specialized tool and not Excel (a database,
   specialized clinical data management system, REDCap etc.)

## And if you use Excel...

 If you use Excel, follow these rules:

   * Versionize your files
   * Check data for consistency (e.g. with R!)
   * Use preselected values 
   * Add a meta-data sheet describing the fields
   * do not color-code or format-code your information (e.g. denoting treatment in bold)
   * Add comments in separate fields
   * NEVER throw away or replace any identifiers, always keep the old identifiers
   * understand what you describe (patients? samples? assays?)
   * understand the difference between wide and long format

## And if you use Excel and don't follow these rules

  * brace yourself for bad times
  * get used to nasty remarks


