---
title: "R crash course / 04"
author: ""
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    widescreen: true
    smaller: true
    css: files/style.css
    logo: files/logo_bihealth_en.png
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
knitr::opts_knit$set(root.dir = "../")
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
library(readxl)
library(beeswarm)
library(tidyr)
library(vioplot)
library(sm)
library(zoo)
library(RColorBrewer)
```

# Matching data sets

## Finding common IDs

We can use `intersect()` for that.

```{r}
set1 <- letters[1:5]
set2 <- letters[3:7]
intersect(set1, set2)
```


## Finding common IDs cont.


```{r}
df1 <- data.frame(ID=sample(letters, 15), value1=rnorm(15))
df2 <- data.frame(ID=sample(letters, 15), value2=rnorm(15))
common <- intersect(df1$ID, df2$ID)

df1[ match(common, df1$ID), ]
df2[ match(common, df2$ID), ]

cbind(df1[ match(common, df1$ID), ], df2[ match(common, df2$ID), ])
```

## Merging

A better, but more complex solution to merge two data frames is the
`merge()` function.

Some SQL terminology:

 * inner join: we are merging only common elements
 * outer join: we are making sure that elements from both sets are taken
   over
 * left outer join: same, but only for the first set
 * right outer join: same, but only for the second set


## Merging (cont.)

Usage of the `merge()` function:

```{r}
merge(df1, df2) # inner join by default
merge(df1, df2, all.x=TRUE) # left outer join
merge(df1, df2, all.y=TRUE) # right outer join
merge(df1, df2, all=TRUE) # outer join
```

Note: tidyverse contains functions `inner_join`, `outer_join` etc. which
look slightly more specific (but do the same).

## Merging (cont.)

The `by`, `by.x` and `by.y` parameters tell `merge` which columns contain
the identifiers by which to join. It is always safer to specify this
parameter!

```{r}
df1 <- data.frame(ID=c("a", "a", "b", "c"), no=c(1, 2, 3, 4), 
  value=rnorm(4))
df2 <- data.frame(ID=c("a", "a", "b", "c"), no=c(1, 2, 3, 4), 
  value=rnorm(4))
merge(df1, df2)           ## not what we wanted
merge(df1, df2, by="ID")  ## not what we wanted
merge(df1, df2, by=c("ID", "no"))
```

## Merging (cont.)

A simpler (but requiring more steps) alternative would be to make a new
identifier:

```{r}
df1$ID2 <- paste(df1$ID, df1$no, sep=".")
df2$ID2 <- paste(df2$ID, df2$no, sep=".")
merge(df1, df2, by="ID2")
```

## Another example

```{r}
df1 <- data.frame(subject.id=1:3, sex=c("M", "F", "F"), 
  age=c(28, 35, 29))
df2 <- data.frame(subject.id=rep(1:3, each=2), 
  timepoint=rep(1:2, 3), value=rnorm(6))
merge(df1, df2, by="subject.id")
```


## Excercise 4.1

The files `expression_data_vaccination_example.xlsx` and
`labresults_full.csv` contain data from the same study.

 1. read the first sheet from the XLSX file and the CSV file.
 3. Which columns ID the subjects? Are there any subjects in common? How do
    you match the subjects?
 5. We are interested only in the following information: Subject ID, ARM
    (group), Time point, sex, age, test name and the actual measurement.
    Are the measurements numeric? Remember, you can use expressions like `[ , c("ARM", "sex") ]` 
    to select the desired columns from a data set.
 4. Use the subjects to merge the two data frames however you see fit. Note
    that there are multiple time points per subject and multiple
    measurements per subject and time point.

# Creating the LABRES data set

## Labres data set

The following slides are for your reference only. You can trace how I
created the `labwide` data frame using meta-data from
`expression_data_vaccination_example.xlsx` and `labresults_full.csv` data
files. The created result can be found in the file
`labresults_wide_full.csv`.

The data set contains laboratory measurements for several parameters (such
as white blood cell count or C-reactive protein) for vaccinees vaccinated
with either Fluad (adjuvant-containing influenza vaccine), Agrippal (same,
but without adjuvant) or placebo, for several days before and after
vaccination.

## Reading and column selection

```{r,eval=TRUE}
targets <- read_excel("Datasets/expression_data_vaccination_example.xlsx", sheet = 1)
targets <- targets[ , c("USUBJID", "Timepoint", "ARM", "SEX", "AGE") ]
lab <- read.csv("Datasets/labresults_full.csv")
lab <- lab[ , 
  c("USUBJID", "LBTEST", "LBTESTCD", "LBORRES", "Timepoint") ]

## remove the "SCREEN" time point
lab <- lab[ lab$Timepoint != "SCREEN", ]

## not what we want!
merged <- merge(targets, lab, by="USUBJID")
table(merged$Timepoint.x, merged$Timepoint.y)
```

## Reading and column selection

We need to choose: do we want to keep all measurements and just fill out
meta-information from `targets`, or do we want only to keep the time points
present in the expression data set?

```{r,eval=TRUE}
## to keep all time points
targets2 <- targets[ , c("USUBJID", "ARM", "SEX", "AGE") ]

## remove duplicated lines
targets2 <- targets2[ !duplicated(targets2$USUBJID), ] 

## merge
merged_full <- merge(targets2, lab, by="USUBJID")

## this one drops all timepoints not present in "targets"
merged <- merge(targets, lab, by=c("USUBJID", "Timepoint"))
```

## Converting to wide format

```{r}
## this does not work as expected, because "LBTESTCD" 
## and "LBTEST" contain the same information – we cannot 
## store both in the column names

labwide <- pivot_wider(merged, names_from="LBTESTCD", values_from="LBORRES")
```

```{r eval=TRUE}
## this works
labwide <- pivot_wider(merged[ , colnames(merged) != "LBTEST"], names_from="LBTESTCD", values_from="LBORRES")
```

## Cleaning up

```{r,eval=TRUE}
sapply(labwide, class) # all measurements are character!
startpos <- which(colnames(labwide) == "ACA")
labwide2 <- data.frame(labwide[,1:(startpos-1)], data.matrix(labwide[,startpos:ncol(labwide)]))
which(is.na(labwide2), arr.ind=T)[,2]

## impute low measurements
labwide[,14] <- gsub("<", "", labwide[,14])
labwide[,16] <- gsub("<", "", labwide[,16])

## these column is all NA
labwide <- labwide[,-17]
labwide <- data.frame(labwide[,1:5], data.matrix(labwide[,6:35]))
```


# Simple statistics with R

## Descriptive statistics

`tapply(x, groups, func)` applies a function `func` to some data `x`
grouped by `groups`.

```{r}
labwide <- read_csv("Datasets/labresults_wide_full.csv")[, -1 ]
mean(labwide$NEUT)
sd(labwide$NEUT)

tapply(labwide$NEUT, labwide$ARM, mean)
tapply(labwide$NEUT, labwide$Timepoint, mean)
tapply(labwide$NEUT, 
  paste(labwide$ARM, labwide$Timepoint, sep="."), mean)
```

## Simple t.test

We define logical vectors to help us select the desired data.

```{r,eval=TRUE}
fluad    <- labwide$ARM == "FLUAD"
agrippal <- labwide$ARM == "AGRIPPAL"

d0 <- labwide$Timepoint == "D0"
d1 <- labwide$Timepoint == "D1"
```

## Simple t.test

Paired or not paired?

```{r,eval=TRUE}
tst <- "NEUT" # lymphocyte count
x <- as.data.frame(labwide)[,tst]
t.test(x[d0 & fluad], x[d1 & fluad])
t.test(x[d0 & fluad], x[d1 & fluad], paired=TRUE)
```

## Broom: cleaning up data

`tidy()` cleans up messy results of various standard statistical tests and
formats them into a nice, clean, TIDY data frame (or, rather, a tibble).

```{r,eval=TRUE,results="markdown"}
library(broom)
tidy(t.test(x[d0 & fluad], x[d1 & fluad], paired=TRUE))
```

## Caveat: paired data

We must make sure that all the data points are paired!

## Excercise 4.2

 * Run t-test for CRP (C-reactive protein), LYM (lymphocytes)
 * `wilcox.test()` is the U test / Mann Whitney test / Wilcoxon test. Try it. Are the p-values different?
 * use `tidy()` to clean up data
 * save the results for CRP and LYM in the data frames (with help of
   `tidy()`).
 * How can you combine two data frames (or tibbles, or matrices) row-wise?
   Look up the `rbind()` function. Create a single data frame with results
   for the CRP and LYM.
 * Adjust the p-values for multiple testing; create a q-value column using
   the function `p.adjust()`. What is the default method for correction for
   multiple testing?


## Simple basic plots (demonstration)

We have two main factors: ARM (the group) and Timepoint (before or 1 day
after vaccination). Therefore, we should plot accordingly.

```{r}
labwide$armcd <- gsub("(.).*", "\\1", labwide$ARM)
labwide$group <- paste(labwide$armcd, labwide$Timepoint, sep=".")

boxplot(NEUT ~ group, data=labwide)
```

## Simple basic plots (demonstration)

```{r,eval=TRUE,echo=FALSE}
labwide$armcd <- gsub("(.).*", "\\1", labwide$ARM)
labwide$group <- paste(labwide$armcd, labwide$Timepoint, sep=".")

boxplot(NEUT ~ group, data=labwide)
```

## Simple basic plots (demonstration)

```{r,eval=TRUE}
library(beeswarm)
beeswarm(NEUT ~ group, data=labwide)
```

## Excercise 4.3

 * Create box plots for selected variables

## ggplot2 vs basic R plots


Basic R plots:

 * super simple for a beginner
 * messy if you try to make it nice
 * you can do anything
 * some tools only produce basic R
 * SUPER fast

ggplot2:

 * produces nice and complex graphics with less effort
 * requires a data frame
 * more uniform
 * very different syntax
 * some things really hard to achieve



## Plot for pairwise data

```{r,eval=TRUE,fig.width=10,fig.height=4.5}
library(ggplot2)
ggplot(labwide, aes(x=Timepoint, y=NEUT, group=USUBJID, color=ARM)) + geom_line() + geom_point()
```

## Making it nicer

```{r,eval=TRUE,fig.width=10,fig.height=4.5}
ggplot(labwide, aes(x=Timepoint, y=NEUT, group=USUBJID)) + geom_line() + geom_point() + 
  facet_grid(. ~ ARM)
```

## Making it nicer

```{r, eval=FALSE}
ggplot(labwide, aes(x=Timepoint, y=NEUT, group=USUBJID)) + 
  geom_line(color="grey") + 
  geom_point() + 
  facet_grid(. ~ ARM) +
  xlab("Day") +
  ylab("Neutrophiles") +
  theme_bw() +
  ggtitle("Change in neutrophile count in different study arms")
```


## Making it nicer

```{r, eval=TRUE,echo=FALSE,fig.width=10,fig.height=5}
ggplot(labwide, aes(x=Timepoint, y=NEUT, group=USUBJID)) + 
  geom_line(color="grey") + 
  geom_point() + 
  facet_grid(. ~ ARM) +
  xlab("Day") +
  ylab("Neutrophiles") +
  theme_bw() +
  ggtitle("Change in neutrophile count in different study arms")
```

