---
title: "R crash course / 04"
author: ""
date: "`r Sys.Date()`"
output:
  ioslides_presentation: default
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
library(readxl)
library(beeswarm)
library(tidyr)
```

# Creating the LABRES data set

## Reading and column selection

```{r,eval=TRUE}
targets <- read_xlsx("Datasets/expression_data_vaccination_example.xlsx", 
  sheet=1)
targets <- targets[ , c("USUBJID", "Timepoint", "ARM", "SEX", "AGE") ]
lab <- read.csv("Datasets/labresults_full.csv")
lab <- lab[ , 
  c("USUBJID", "LBTEST", "LBTESTCD", "LBORRES", "Timepoint") ]

## remove the "SCREEN" time point
lab <- lab[ lab$Timepoint != "SCREEN", ]

## not what we want!
merged <- merge(targets, lab, by="USUBJID")
table(merged$Timepoint.x, merged$Timepoint.y)
```

## Reading and column selection

We need to choose: do we want to keep all measurements and just fill out
meta-information from `targets`, or do we want only to keep the time points
present in the expression data set?

```{r,eval=TRUE}
## to keep all time points
targets2 <- targets[ , c("USUBJID", "ARM", "SEX", "AGE") ]

## remove duplicated lines
targets2 <- targets2[ !duplicated(targets2$USUBJID), ] 

## merge
merged_full <- merge(targets2, lab, by="USUBJID")

## this one drops all timepoints not present in "targets"
merged <- merge(targets, lab, by=c("USUBJID", "Timepoint"))
```

## Converting to wide format

```{r}
## this does not work as expected, because "LBTESTCD" 
## and "LBTEST" contain the same information – we cannot 
## store both in the column names

labwide <- spread(merged, key="LBTESTCD", value="LBORRES")
```

```{r eval=TRUE}
## this works
labwide <- spread(merged[ , colnames(merged) != "LBTEST"], key="LBTESTCD", value="LBORRES")
```

## Cleaning up

```{r,eval=TRUE}
sapply(labwide, class) # all measurements are character!
startpos <- which(colnames(labwide) == "ACA")
labwide2 <- data.frame(labwide[,1:(startpos-1)], data.matrix(labwide[,startpos:ncol(labwide)]))
which(is.na(labwide2), arr.ind=T)[,2]

## impute low measurements
labwide[,14] <- gsub("<", "", labwide[,14])
labwide[,16] <- gsub("<", "", labwide[,16])

## these column is all NA
labwide <- labwide[,-17]
labwide <- data.frame(labwide[,1:5], data.matrix(labwide[,6:35]))
```


# Simple statistics with R

## Descriptive statistics

```{r}
mean(labwide$NEUT)
sd(labwide$NEUT)

tapply(labwide$NEUT, labwide$ARM, mean)
tapply(labwide$NEUT, labwide$Timepoint, mean)
tapply(labwide$NEUT, 
  paste(labwide$ARM, labwide$Timepoint, sep="."), mean)
```

## Simple t.test

```{r,eval=TRUE}
fluad <- labwide$ARM == "FLUAD"
agrippal <- labwide$ARM == "AGRIPPAL"

d0 <- labwide$Timepoint == "D0"
d1 <- labwide$Timepoint == "D1"
```

## Simple t.test

Paired or not paired?

```{r,eval=TRUE}
tst <- "NEUT" # lymphocyte count
x <- labwide[ ,tst]
t.test(x[d0 & fluad], x[d1 & fluad])
t.test(x[d0 & fluad], x[d1 & fluad], paired=TRUE)
```

## Caveat: paired data

We must make sure that all the data points are paired!

## Excercise 4.1

 * Run t-test for CRP (C-reactive protein), LYM (lymphocytes)
 * `wilcox.test()` is the U test / Mann Whitney test / Wilcoxon test. Try it


## Basic plots (demonstration)

We have two main factors: ARM (the group) and Timepoint (before or 1 day
after vaccination). Therefore, we should plot accordingly.

```{r}
armcd <- gsub("(.).*", "\\1", labwide$ARM)
group <- paste(armcd, labwide$Timepoint, sep=".")

x <- labwide[ , tst]
boxplot(x ~ group)
```

## Basic plots (demonstration)

```{r,eval=TRUE,echo=FALSE}
armcd <- gsub("(.).*", "\\1", labwide$ARM)
group <- paste(armcd, labwide$Timepoint, sep=".")

x <- labwide[ , tst]
boxplot(x ~ group)
```

## Basic plots (demonstration)

```{r,eval=TRUE}
beeswarm(x ~ group)
```

## Excercise 4.2

 * Create box plots for selected variables

## Plotting principles

 * plotting parameters
 * `xlab`, `ylab`, `main`

## Plot for pairwise data

```{r,eval=FALSE}
x <- as.numeric(factor(group))
y <- labwide[ , tst]

plot(x, y, xaxt="n", bty="n", pch=19, col="grey",
  xlab="Vaccine", ylab="Lymphocyte count")
segments(x[d0], y[d0], x[d1], y[d1])
axis(1, at=2*(1:3)-.5, labels=levels(factor(labwide$ARM)))
```

## Plot for pairwise data

```{r,eval=TRUE,echo=FALSE}
x <- as.numeric(factor(group))
y <- labwide[ , tst]

plot(x, y, xaxt="n", bty="n", pch=19, col="grey")
segments(x[d0], y[d0], x[d1], y[d1])
axis(1, at=2*(1:3)-.5, labels=levels(factor(labwide$ARM)))
```



# Multivariate data

## PCA

 * An example of a complex object type
 * `pch`, `bty`, `lty`

```{r}
data(iris)
pca <- prcomp(iris[,1:4], scale.=TRUE)
plot(pca$x[,1], pca$x[,2], col=factor(iris$Species),
  pch=19, bty="n")
legend("topright", levels(factor(iris$Species)),
  pch=19, col=1:3, bty="n")
```

## PCA (cont.)

```{r,eval=TRUE,echo=FALSE}
data(iris)
pca <- prcomp(iris[,1:4], scale.=TRUE)
plot(pca$x[,1], pca$x[,2], col=factor(iris$Species),
  pch=19, bty="n")
legend("topright", levels(factor(iris$Species)),
  pch=19, col=1:3, bty="n")
```


## Farbenlehre
 
 * numeric values (simple colors)
 * color names: see http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
 * (R, G, B) color specification
    * Hexadecimal notation
    * Basic colors:
        * `#FF0000` (red)
        * `#00FF00` (green)
        * `#0000FF` (blue)
  * friendly color palettes: `RColorBrewer` package


## RColorBrewer

```{r}
library(RColorBrewer)
display.brewer.all()
pal <- brewer.pal(n=3, name="Dark2")

cols <- pal[ factor(iris$Species) ]
plot(pca$x[,1], pca$x[,2], col=cols,
  pch=19, bty="n")
legend("topright", levels(factor(iris$Species)),
  pch=19, col=pal, bty="n")
```


## RColorBrewer (cont.)

```{r,eval=TRUE,echo=FALSE}
library(RColorBrewer)
pal <- brewer.pal(n=3, name="Dark2")

cols <- pal[ factor(iris$Species) ]
plot(pca$x[,1], pca$x[,2], col=cols,
  pch=19, bty="n")
legend("topright", levels(factor(iris$Species)),
  pch=19, col=pal, bty="n")
```


## Putting two plots on one graphics

```{r}
par(mfrow=c(1,2))
plot(pca$x[,1], pca$x[,2], col=cols, pch=19, bty="n")
plot(pca$x[,3], pca$x[,4], col=cols, pch=19, bty="n")
legend("topright", levels(factor(iris$Species)),
  pch=19, col=pal, bty="n")
```

## Putting two plots on one graphics (cont.)

```{r,eval=TRUE,echo=FALSE,fig.width=8,fig.height=4}
par(mfrow=c(1,2))
plot(pca$x[,1], pca$x[,2], col=cols, pch=19, bty="n")
plot(pca$x[,3], pca$x[,4], col=cols, pch=19, bty="n")
legend("topright", levels(factor(iris$Species)),
  pch=19, col=pal, bty="n")
```

## Excercise 

 * Compute and display a PCA for the labwide data set. Which components
   appear to be related to the different sample groups? 

## Playing with PCA: Gene expression

(demonstration)

## The mtcars data set – correlations & co.


```{r}
data(mtcars)
pairs(mtcars[,1:5])
cor(mtcars$hp, mtcars$mpg)
cor.test(mtcars$hp, mtcars$mpg)
cor.test(mtcars$hp, mtcars$mpg, method="spearman")
```

## mtcars and PCA

```{r}
pca <- prcomp(mtcars, scale.=TRUE)
cor(pca$x[,1], mtcars[,1])
```


## Linear regression

```{r}
mod <- lm(mpg ~ hp + disp)
summary(mod)
```


