---
title: "R crash course / 02"
author: ""
date: "`r Sys.Date()`"
output:
  ioslides_presentation: default
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=TRUE,autodep=TRUE, results="hide")
```

## Data matrices

Much like vectors, matrices can only hold one data type (e.g. only numeric
or only character).

```{r}
m <- matrix(1:18, ncol=3, nrow=6)
# compare with
m <- matrix(1:18, ncol=3, nrow=6, byrow=TRUE)
dim(m)
ncol(m)
nrow(m)
```

## Accessing columns, rows and element

`matrix[row, column]`

So, for example:

```{r}
m[1, ] # vector which is the first row
m[, 2] # vector which is the first column
m[3, 1] # first element of the third row
```

## Row and column names

We can name rows and columns of a matrix and use the names to access the
rows and columns:

```{r}
colnames(m) <- letters[1:ncol(m)]
rownames(m) <- LETTERS[1:nrow(m)]

m["A", "b"]
m["B", ] # etc.
```

## Remember

 * If you select a single column or a single row, you will get a vector
 * If you select more than one row or column, you will get a (smaller) matrix
 * If you select more rows or columns than are present, you will get a
   "subscript out of bonds" error

## Demonstration


## Data frames

Data frames are a bit like matrices, but every column can store different
type of data. In this, they are more like lists (which they in fact are).

```{r}
names <- c("January", "Bill")
lastn <- c("Weiner", "Gates")
age   <- c(1001, 64)

d <- data.frame(names=names, last_names=lastn, age=age)
class(d)
class(d[,1])
class(d[,3])
```

## Accessing elements in data frames

You can access the data frame elements much like the elements of a matrix.

However, since data frames are lists, the list operator (`$`) also works:

```{r}
d$names # same as d[,1] or d[, "names"]
d$lastn
d$lastn[1]
```

However, note that when you select a row, you will get a data frame, not a
vector. This is because each of the column can be of different type, and
vectors can hold only one type of data.

## Matrices vs data frames

 Caveats:

  * data frames sometimes turn strings to factors (more on that later), which may have really
    disastrous consequences, use `stringsAsFactors=FALSE`
  * small typos can turn a numeric vector into a string
  * factors are dangerous to work with, use them cautiously

  Gory details: matrices are a basic data type. Data frames are a list.


## Excercise 2.1

 * Create a 5x3 matrix with random numbers. Use `matrix` and `rnorm`.
 * Turn the matrix into a data frame. Use `as.data.frame` for that.
 * Add column and row names.
 * Add a column. Each value in the column should be "A" (a string). Use the
   `rep` function for that.
 * Add a column with numbers 1 to 5. Use the `seq` function for that.

## Using rep

`rep()` is used to replicate vectors.

```{r}
rep(c("A", "B"), 5) 
# result:
#  [1] "A" "B" "A" "B" "A" "B" "A" "B" "A" "B"

rep(c("A", "B"), each=5)
# result
# [1] "A" "A" "A" "A" "A" "B" "B" "B" "B" "B"
```

# Reading and writing data

## Reading data

Main data types you will encounter:

---------------------------- ------------------------------- --------------------------
Data type                    Function                        Notes
---------------------------- ------------------------------- --------------------------
TSV / TAB separated values   `read.table()`                  Tab is `\t`.
                                                             A very general and 
                                                             customizable function

CSV / comma separated        `read.csv()`                    a wrapper around
                                                             `read.table()`

XLS (old Excel)              `read_xls()`                    Just don't use it.
                                                             From the
                                                             `readxl`
                                                             package.

XLSX (new Excel)             `read_xlsx()`                   From the `readxl` package.
                                                             You need to
                                                             provide the
                                                             sheet number
                                                             you wish to
                                                             read. 
                                                             Note: returns a tibble,
                                                             not a data frame! 
---------------------------- ------------------------------- --------------------------



## Examples (demonstration)

 * `header=TRUE`
 * `colnames`
 * `rownames`
 * `sep=","`
 * `stringsAsFactors`
 * `quote`
 * `comment.char`

## Tibbles / readxl

tibbles belong to the tidyverse. They are nice to work with and very
useful, but we will stick to data frames in this course. Therefore, do

```{r}
mydataframe <- as.data.frame(read_xlsx("file.xlsx"))
```

One crucial difference between tibble and data frame is that `tibble[ , 1 ]` 
returns a tibble, while `dataframe[ , 1]` returns a vector. The second
crucial difference is that it does not support row names (on purpose!).

## Cleaning up data

`unique` returns the unique values in a vector. Use it to check the
correctness of the columns.

Use `class()` to make sure that the columns are what you think they are.

Use `summary()` to gain an overview of the data.

Convert with `factor` when necessary (and only then).

## Substitutions

```{r}
group <- c("Control", " control", "control ", "Control   ")
group <- gsub("Control", "control", group)
group <- gsub(" ", "", group)
```

## Regular expressions

```{r}
group <- c("Control", " control", "control ", "Control   ")

# substitute at the end of the string
gsub(" $", "", group)
# at the beginning
gsub("^ ", "", group)

# any number of occurences
gsub(" *", "", group)
```

## Regular expression cont.

```{r}
foo <- c("a1_123", "a1_231", "a2_17", "b2_23", "c2_889")

# substitute digits only
gsub("[0-9]*", "", foo)

# substitute letters only
gsub("[a-z]*", "", foo)

# substitute everything after "_"
gsub("_.*", "", foo)
```

## Writing data

Main data types you will encounter:

---------------------------- ------------------------------- --------------------------
Data type                    Function                        Notes
---------------------------- ------------------------------- --------------------------
TSV / TAB separated values   `write.table()`                 Tab is `\t`.
                                                             A very general and 
                                                             customizable function
                                                             Warning:
                                                             column names 

CSV / comma separated        `write.csv()`                    a wrapper around
                                                             `read.table()`

XLS (old Excel)                                              Just don't use it. No, 
                                                             seriously, don't.

XLSX (new Excel)             `write_xlsx()`                  From the `writexl` 
                                                             package.
                                                             Warning:
                                                             row names
                                                             omitted!
---------------------------- ------------------------------- --------------------------


## Demonstration


## Excercise 2.2

Read, inspect and correct (if reasonable) the following files:

 * `cars.xlsx`
 * `iris.csv`
 * `expression_data_vaccination_example.xlsx` (large!)

The function `readxl_example("deaths.xls")` returns a file name. Read this
file. How can you omit the lines at the top and at the bottom of the file?
(hint: `?read_xls`). How can you force the date columns to be interpreted
as dates and not numbers?


## Wide and Long format (demonstration)


## Converting from wide to long:

```{r}
wide <- read.table(header=TRUE, text='
 subject sex control cond1 cond2
       1   M     7.9  12.3  10.7
       2   F     6.3  10.6  11.1
       3   F     9.5  13.1  13.8
       4   M    11.5  13.4  12.9
')
tidyr::gather(wide, key="condition", 
             value="measurement",  control:cond2)
```

`tidyr::gather` means we are accessing the `gather()` function from the
`tidyr` package without having to call `library(tidyr)` first.

## Converting from long to wide

```{r}
long <- read.table(header=TRUE, text='
 subject  sampleID sex condition measurement
       1  ID000001 M   control         7.9
       1  ID000002 M     cond1        12.3
       1  ID000003 M     cond2        10.7
       2  ID000004 F   control         6.3
       2  ID000005 F     cond1        10.6
       2  ID000006 F     cond2        11.1
       3  ID000007 F   control         9.5
       3  ID000008 F     cond1        13.1
       3  ID000009 F     cond2        13.8
')
```

## Converting from long to wide

```{r}
## not what we wanted!!!
tidyr::spread(long, key="condition", value="measurement")

## Instead: 
long <- long[ , c("subject", "condition", "measurement") ]
tidyr::spread(long, key="condition", value="measurement")
```

## Excercise 2.3

Convert the following files to long format:

 * `iris.csv`
 * `mtcars_wide.csv`
 * `labresults_wide.csv`
 * `cars.xlsx` (tricky!)

## Excercise 2.4

For the first sheet ("targets") from the file
`expression_data_vaccination_example.xlsx`, create a new table. For each
Patient ID there should be one row. The row should contain a column for
each day (D0, D1 etc.) containing the sample IDs collected on that day.


