---
title: "Exercises"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The lily pond problem

There is a lake in a garden. Every day, the water lilies cover twice as much area as the previous day. On the first day, the water lilies cover 1/100th of the area of the lake.

Formula for day n:

$$A(n) = A_0 \cdot 2^n$$

In the following, x are the days and y is the fraction of pond covered on day n. We start at day 0, when the coverage is 1/100th.

```{r echo=TRUE}
frac_init <- 1/100

x <- 1:7

y <- frac_init * 2^x
plot(x, y, type="l", lty=3, bty="n", 
     xlab="Day", ylab="Fraction of pond covered", 
     main="Lily pond", col="red", log="y")
y2 <- frac_init * 1.5^x
lines(x, y2, col="blue", lty=2)
abline(h=.5, col="grey")
```

And here, the ggplot2 version (you need nota worry about it)

```{r echo=FALSE}
library(ggplot2)
df <- data.frame(x=x, y=y)
ggplot(df, aes(x=x, y=y)) + geom_line() + theme_bw() + xlab("Day") + ylab("Fraction of pond covered") +
  ggtitle("Lily pond") + geom_hline(yintercept =.5, color="grey")
```


# Dummy t test


```{r}
weight <- c(16, 17, 21, 23, 15)
fac <- factor(c("male", "male", "female", "female", "male"))
res <- t.test(weight ~ fac)
```

The p-value from the t-test was `r format.pval(res$p.value, digits=2)`.
